Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2017-12-15 article.tex
Cambios y ajustes sobre imagenes. Version base

\firstnote{.}

@@

% Keywords
\keyword{EEG; BCI; P300; ALS; NBNN; HOG; SIFT}

@@

In~\citep{Ramele2016},  authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves~\citep{Ben-Simon2013}  and Motor Imagery rolandic central $\mu$ rhythms~\citep{Hartman2005} using gradient histograms orientations. Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  Its validity is verified by processing offline data for ALS patients and for healthy subjects. 
The proposed approach is based on the morphological analysis of the shape of the EEG signal~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and it was built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots~\citep{Hartman2005}.

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of oriented gradients extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradients of the Signal Plot method is explained. Section~\ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

@@

\subsection{Feature Extraction based on Histogram of Gradients of the Signal Plot} \label{Feature}

The underlying idea of this method is to generate templates for the signal \citep{Ramele2016}, based on the image plot.  Hence, the first step is the signal transformation into a temporary binary image.

@@

\subsubsection{Feature Extraction: Histogram of Oriented Gradients}
\label{SIFT}

@@

@@

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph2.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}

@@

EEG data is collected in a single recording session. Participants are seated in a comfortable chair, with their vision aligned to a computer screen located one meter in front of them.  The handling and processing of the data and stimulations is handled by the OpenVibe platform.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole signal information.  The vertical size $4 \cdot 3 s_y$ of the patch is required to be long enough as to contain the signal $\Delta  \mu V $, which is the peak-to-peak amplitude of the transient event. The horizontal size $ 4 \cdot 3 s_x$ , on the other hand, needs to cover the entire duration in seconds of the transient signal event, $ \lambda $.   }
\label{fig:sampledescriptor2}
\end{figure}

@@

Table~\ref{tab:results} shows the results of applying the algorithm to the subjects of the public dataset. The percentage of correctly spelled letters is calculated while performing an offline BCI Simulation for the public dataset of ALS patients.  From the 7 words for each subject, the first 3 were used as training, and the remaining 4 for testing.  The best performing channel is informed as well. The chance level is $2\%$.  
%JG esto después lo hablamos, ¿porque lo pondrías aca?
%DRAFT: Traditionally P300 can be seen in this channel and this other channel.  However, Wolpaw showed that the response may also arise in occipital channels as well.  The authors of the original paper concluded that Cz was the best performing channel, more representative of the difference in amplitudes. However, in our approach occipital channels,  PO8 and PO7 showed higher performances \citep{Huggins2016,Jure2016}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{performance.eps}
\caption{Performance curves for the eight subjects included in the published dataset.  Three out of eight subjects achieved the necessary performance to implement a valid P300 speller.}
\label{fig:performance}
\end{figure}

@@

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each healthy subject. Chance level is $2\%$. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{cccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &     0.7000  &  0.3500  \\
1     &     Cz   &    0.1000  &  0.0500 \\
1     &     Cz   &     0.1000  &  0.1000 \\
1     &     Cz   &     0.1000  &  0.1500 \\
1     &     Cz   &     0.1000  &  0.1000 \\
1     &     Cz   &     0.5500  &  0.4000 \\
1     &     Cz   &     0.1500   & 0.1000 \\
1     &     Cz   &     0.4500   & 0.2000 \\
1     &     Cz   &     0.3000  &  0.2000 \\
1     &     Cz   &     0.2500  &  0.1500 \\
1     &     Cz   &     0.5500  &  0.3000 \\
1     &     Cz   &     0.3500  &  0.1000 \\
1     &     Cz   &     0.3000  &  0.3500 \\
1     &     Cz   &     0.5000  &  0.4000 \\
1     &     Cz   &     0.3000  &  0.1500 \\
1     &     Cz   &     0.3000  &  0.3000 \\
1     &     Cz   &     0.3500  &  0.3500 \\
1     &     Cz   &          0      & 0.0500 \\
1     &     Cz   &     0.7000   & 0.5000 \\
1     &     Cz   &     0.4000  &  0.3000 \\

@@

\noindent EEG: Electroencephalography\\
BCI: Brain Computer Interfaces\\
SNR: Signal to Noise Ratio\\
CNS: Central Nervous System\\
ALS: Amyotrophic Lateral Sclerosis\\
ERP: Event-Related Potential\\
P300: Positive deflection of an Event-Related Potential which occurs 300 ms after onset of stimulus\\
ITR: Information Transfer Rate\\
BTR: Bit Transfer Rate\\
SIFT: Scale Invariant Feature Transform\\
NBNN: Naive Bayes Nearest Neighbor\\
HOG: Histogram Of Gradients}

>>>>

%\firstnote{.}

@@

% Keywords
\keyword{EEG; BCI; P300; ALS; NBNN; HGOSP; SIFT}

@@

In~\citep{Ramele2016},  authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves~\citep{Ben-Simon2013}  and Motor Imagery rolandic central $\mu$ rhythms~\citep{Hartman2005} using the histogram of gradient orientations of signal plots. Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  Its validity is verified by processing offline data for ALS patients and for healthy subjects. 
The proposed approach is based on the morphological analysis of the shape of the EEG signal~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and it was built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots~\citep{Hartman2005}.

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of gradient orientations extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradient Orientations of the Signal Plot method is explained. Section~\ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

@@

\subsection{Feature Extraction based on Histogram of Gradient Orientations of the Signal Plot} \label{Feature}

The underlying idea of this method is to generate templates for the signal \citep{Ramele2016}, based on the image plot.  Hence, the first step is the processing of the EEG signal.

@@

\subsubsection{Feature Extraction: Histogram of Gradient Orientations}
\label{SIFT}

@@

%$(\hat{r},\hat{c})$

%$d^{col}_{\hat{c}}$          $d^{row}_{\hat{r}}$

@@

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

@@

EEG data is collected in a single recording session. Participants are seated in a comfortable chair, with their vision aligned to a computer screen located one meter in front of them.  The handling and processing of the data and stimulations is handled by the OpenVibe platform~\citep{Renard2010}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole transient event.  The size of the patch is $S_p \times S_p$ pixels. The vertical size consists of $4$ blocks $B_{i,j}$  of size $3 s_y$ pixels which is long enough as to contain the signal $\Delta  \mu V $, the peak-to-peak amplitude of the transient event. The horizontal size on the other hand, includes $4$ blocks  of $3 s_x$ and cover the entire duration in seconds of the transient signal event, $ \lambda $.   }

\label{fig:sampledescriptor2}
\end{figure}

@@

Table~\ref{tab:results} shows the results of applying the algorithm to the subjects of the public dataset of ALS patients. The percentage of correctly spelled letters is calculated while performing an offline BCI Simulation.  From the 7 words for each subject, the first 3 were used as training, and the remaining 4 for testing.  The best performing channel is informed as well. The chance level is $2\%$.  
%JG esto después lo hablamos, ¿porque lo pondrías aca?
%RR: esto corresponde al punto 4.1 que dijo el revisor y tiene que ver con proveer informacion del canal en el que la clasificacion dio mejor.
%DRAFT: Traditionally P300 can be seen in this channel and this other channel.  However, Wolpaw showed that the response may also arise in occipital channels as well.  The authors of the original paper concluded that Cz was the best performing channel, more representative of the difference in amplitudes. However, in our approach occipital channels,  PO8 and PO7 showed higher performances \citep{Huggins2016,Jure2016}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{performance.eps}
\caption{Performance curves for the eight subjects included in the ALS dataset.  Three out of eight subjects achieved the necessary performance to implement a valid P300 speller.}
\label{fig:performance}
\end{figure}

@@

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each healthy subject. Chance level is $2\%$. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{cccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &     0.7000  &  $45\%$  \\
1     &     Cz   &    0.1000   &  $10\%$ \\
1     &     Cz   &     0.1000  &  $15\%$ \\
1     &     Cz   &     0.1000  &  $20\%$  \\
1     &     Cz   &     0.1000  &  $10\%$ \\
1     &     Cz   &     0.5500  &  $25\%$ \\
1     &     Cz   &     0.1500  &  $15\%$ \\
1     &     Cz   &     0.4500  &  $20\%$ \\
1     &     Cz   &     0.3000  &  $20\%$ \\
1     &     Cz   &     0.2500  &  $20\%$ \\
1     &     Cz   &     0.5500  &  $15\%$ \\
1     &     Cz   &     0.3500  &  $30\%$ \\
1     &     Cz   &     0.3000  &  $10\%$ \\
1     &     Cz   &     0.5000  &  $35\%$ \\
1     &     Cz   &     0.3000  &  $45\%$ \\
1     &     Cz   &     0.3000  &  $10\%$ \\
1     &     Cz   &     0.3500  &  $30\%$ \\
1     &     Cz   &          0      &  $45\%$ \\
1     &     Cz   &          0      &  $15\%$ \\
1     &     Cz   &     0.7000  &  $65\%$ \\
1     &     Cz   &     0.4000  &  $40\%$ \\

@@

\noindent EEG: Electroencephalography\\
BCI: Brain Computer Interfaces\\
SNR: Signal to Noise Ratio\\
CNS: Central Nervous System\\
ALS: Amyotrophic Lateral Sclerosis\\
ERP: Event-Related Potential\\
P300: Positive deflection of an Event-Related Potential which occurs 300 ms after onset of stimulus\\
ITR: Information Transfer Rate\\
BTR: Bit Transfer Rate\\
SIFT: Scale Invariant Feature Transform\\
NBNN: Naive Bayes Nearest Neighbor}
